Timid tides or tall tails? \\ Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-05-31 thesis/introduction.tex
thesis updates / reorganization of RV modeling

@@

\includegraphics[width=5.41667in,height=5.41667in]{/Users/daniel/thesis/figures/scl_des_dr2.png}Figure:
Image of the Sculptor dwarf spheroidal galaxy from Dark Energy Survey
Data Release 2 {[}\citet{abbott+2021}; image created with HiPS2FITS{]}.
Sculptor appears as a fairly prominent, extended over density of
predominantly faint, red stars. 0.5 degree field of view centred on
Sculptor.

\includegraphics[width=5.41667in,height=5.41667in]{figures/umi_DSS2_0.75deg.png}Figure:
Image of Ursa Minor dwarf galaxy from the Digitized Sky Survey 2 (0.75
deg field of view tangent plane). UMi appears as a diagonal/elliptical
haze of faint, reddish stars from the top left to the bottom right. Even
as classical dwarf, Ursa minor is fairly diffuse and does not stand
obviously out from the background.

>>>>

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{/Users/daniel/thesis/figures/scl_des_dr2.png}
\caption[Picture of Sculptor]{Image of the Sculptor dwarf spheroidal
galaxy from Dark Energy Survey Data Release 2 {[}\citet{abbott+2021};
image created with HiPS2FITS{]}. Sculptor appears as a fairly prominent,
extended over density of predominantly faint, red stars. 0.5 degree
field of view centred on Sculptor. \textbf{TODO: combine with UMi, maybe
add LMC and Fornax?}}\label{fig:scl_image}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/umi_DSS2_0.75deg.png}
\caption[Picture of Ursa Minor]{Image of Ursa Minor dwarf galaxy from
the Digitized Sky Survey 2 (0.75 deg field of view tangent plane). UMi
appears as a diagonal/elliptical haze of faint, reddish stars from the
top left to the bottom right. Even as classical dwarf, Ursa minor is
fairly diffuse and does not stand obviously out from the
background.}\label{fig:umi_image}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/mw_satellites_ongaia.pdf}}
\caption[Dwarf galaxies sky position]{The location of MW dwarf galaxies
on the sky. Dwarf galaxies are taken as confirmed dwarfs with MW or LMC
hosts from the \citet{pace2024} catalogue (version 1.0.3. We label all
classical dwarfs and the Magellanic clouds, changing symbols based on
the relevance to this work (classical dwarfs, Magellanic clouds REF, and
the systems in question or comparison). \textbf{TODO: use
abbreviations}}
\end{figure}

@@

### 2025-05-31 thesis/methods.tex
thesis updates / reorganization of RV modeling

The 2D exponential stellar profile is given by \begin{equation}
\Sigma_\star(R) = A e^{-R / R_s}
\end{equation} where \(A\) is some normalization and \(R_s\) is the
exponential scale radius. The 3D deprojected profile is found through
abel inversion (e.g.~Rapha\ldots), i.e. \begin{equation}
\rho_\star (r) =- \frac{1}{\pi}\int_r^\infty \frac{d\Sigma}{dR} \frac{1}{\sqrt{R^2 - r^2}} dR  = \frac{\Sigma_0}{\pi R_s^2}\,K_0(r/R_s)
\end{equation} where \(K_0\) is the 0th order modified Bessel function
of the second type.

@@

A Plummer profile is defined by \begin{equation}
\Sigma(R) = \frac{M}{4\pi R_s^2} \left(1 + \frac{R^2}{R_s^2}\right)^{-2} ,
\end{equation} where \(M\) is the total mass and \(R_s\) is the
characteristic scale radius. The density is \begin{equation}
\rho(r) = \frac{3M}{4\pi\,R_s^3} \left(1 + \frac{r^2}{R_s^2}\right)^{-5/2}.
\end{equation}

@@

>>>>

Perhaps the simplest in form, the 2D exponential is commonly used to
describe dwarf galaxy's density profiles as well as stellar disks. The
stellar profile is given by \begin{equation}
\Sigma_\star(R) = A e^{-R / R_s}
\end{equation} where \(A\) is the normalization and \(R_s\) is the
exponential scale radius. The 3D deprojected profile is found through
Abel inversion \citep[e.g.][]{errani+2024}: \begin{equation}
\rho_\star (r) =- \frac{1}{\pi}\int_r^\infty \frac{d\Sigma}{dR} \frac{1}{\sqrt{R^2 - r^2}} dR  = \frac{\Sigma_0}{\pi R_s^2}\,K_0(r/R_s)
\end{equation} where \(K_0\) is the 0th order modified Bessel function
of the second type.

@@

The Plummer density profile is generally considered to be a reasonable
empirical fit to Globular Clusters, where it was first proposed
\citep{sadfjk}. Plummer profiles are also sometimes used to describe
dwarf spheroidal galaxies REFS and galactic bulges REF. A Plummer
profile is defined by \begin{equation}
\Sigma(R) = \frac{M}{4\pi R_s^2} \left(1 + \frac{R^2}{R_s^2}\right)^{-2} ,
\end{equation} where \(M\) is the total mass and \(R_s\) is the
characteristic scale radius. The 3D density is \begin{equation}
\rho(r) = \frac{3M}{4\pi\,R_s^3} \left(1 + \frac{r^2}{R_s^2}\right)^{-5/2}.
\end{equation}

@@

\section{Appendix: Numerical convergence and
parameters}\label{appendix-numerical-convergence-and-parameters}

Here, we describe some convergence tests to ensure our methods and
results are minimally impacted by numerical limitations and assumptions.
See \citet{power+2003} for a detailed discussion of various assumptions
and parameters used in N-body simulations.

\subsubsection{Softening}\label{softening}

\citet{power+2003} suggest the empirical rule that the ideal softening
(balancing integration time and only compromising resolution in
collisional regime) is \begin{equation}
h_{grav} = 4 \frac{R_{200}}{\sqrt{N_{200}}}
\end{equation}

For our isolation halo (\(M_s=2.7\), \(r_s=2.76\)) and with \(10^7\)
particles, this works out to be \(0.044\,{\rm kpc}\).We adpoted the
slightly smaller softening which was reduced by a factor of
\(\sqrt{10}\) which appears to improve agreement slightly in the
innermost regions. However, this choice likely unnecessarily increases
computation time for the relative gain in accuracy.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/iso_converg_softening.png}}
\caption{Softening convergence}\label{fig:softening_convergence}
\end{figure}

\subsubsection{Timestepping and force
accuracy}\label{timestepping-and-force-accuracy}

In general, we use adaptive timestepping and relative opening criteria
for gravitational force computations. To verify that these choices and
associated accuracy parameters minimally impact convergence or speed, we
show a few more isolation runs (using only 1e5 particles)

\begin{itemize}
\tightlist
\item
  constant timestep (\ldots), approximantly half of minimum timestep
  with adaptive timestepping
\item
  geometric opening, with \(\theta = 0.5\).
\item
  strict integration accuracy, (facc = \ldots.)
\end{itemize}

\subsubsection{Alternative methods}\label{alternative-methods}

\begin{itemize}
\tightlist
\item
  FMM
\item
  PMM-tree
\item
  Gadget2
\item
  etc.
\end{itemize}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/iso_converg_methods.png}}
\caption{Isolation method convergence}\label{fig:methods_convergence}
\end{figure}

\subsubsection{Fiducial Parameters}\label{fiducial-parameters}

Note that we use code units which assume that \(G=1\) for convenience
and numerical stability. The conversion between code units to physical
units is (for our convention):

\begin{itemize}
\tightlist
\item
  1 length = 1 kpc
\item
  1 mass unit = \(10^{10}\) Msun
\item
  1 velocity unit = 207.4 km/s
\item
  1 time unit = 4.715 Myr
\end{itemize}

Most parameters below are not too relevant or have been discussed or are
merely dealing with cpu and IO details. The changes between simulation
runs primarily affect the integration time, output frequency, and
softening. Otherwise, we leave all other parameters fixed.

\begin{verbatim}
#======IO parameters======

#---Filenames
InitCondFile                initial
OutputDir                   ./out
SnapshotFileBase            snapshot
OutputListFilename          outputs.txt

#---File formats 
ICFormat                    3       # use HDF5
SnapFormat                  3

#---Mem & CPU limits
TimeLimitCPU                86400
CpuTimeBetRestartFile       7200
MaxMemSize                  2400

#---Time
TimeBegin                   0
TimeMax                    2120      # 10 Gyr

#---Output frequency
OutputListOn                0
TimeBetSnapshot             10
TimeOfFirstSnapshot         0 
TimeBetStatistics           10
NumFilesPerSnapshot         1
MaxFilesWithConcurrentIO    1

#=======Gravity======

#---Timestep accuracy
ErrTolIntAccuracy           0.01
CourantFac                  0.1     # ignored; for SPH
MaxSizeTimestep             0.5
MinSizeTimestep             0.0

#---Tree algorithm
TypeOfOpeningCriterion      1       # Relative
ErrTolTheta                 0.5     # mostly used for Barnes-Hut
ErrTolThetaMax              1.0     # (used only for relative)
ErrTolForceAcc              0.005   # (used only for relative)

#---Domain decomposition: should only affect performance
TopNodeFactor                       3.0
ActivePartFracForNewDomainDecomp    0.02

#---Gravitational Softening
SofteningComovingClass0      0.044  # HALO dependent
SofteningMaxPhysClass0       0      # ignored; for cosmological
SofteningClassOfPartType0    0
SofteningClassOfPartType1    0

#=======Miscellanius=======

# probably do not need to change the options below

#---Unit System
UnitLength_in_cm            1 
UnitMass_in_g               1
UnitVelocity_in_cm_per_s    1 
GravityConstantInternal     1

#---Cosmological Parameters 
ComovingIntegrationOn      0 # no cosmology
Omega0                     0
OmegaLambda                 0 
OmegaBaryon                 0
HubbleParam                 1
Hubble                      100
BoxSize                     0

#---SPH
ArtBulkViscConst             0.8
MinEgySpec                   0
InitGasTemp                  100

#---Initial density estimate (SPH)
DesNumNgb                   64
MaxNumNgbDeviation          1
\end{verbatim}
